MSstatsPTM statistical relative quantification of post-translational modifications in global proteomics experiments

### 2022-02-17 main/MSstatsPTM_Main.tex
intro changes

This manuscript proposes a general statistical analysis framework that detects relative changes in post-translational modifications. The proposed framework combines the quantitative information pertaining to a PTM site over peptides and replicates in multiple conditions, and takes into account various sources of variation and confounding. The framework requires an experimental design that, for each sample, quantifies the peptides with post-translational modifications, as well as  peptides from the same proteins with no modification sites. It is applicable to both label-free and tandem mass tag (TMT)-based acquisitions. The proposed framework summarizes the abundances of peptides with and without the modification sites, and fits a linear mixed effects model that automatically adjusts to the biological and technological aspects of the experimental design. \todo{For questions focusing on PTM, can analyze changes in the modified peptides. For eliminating the confounding, combine...} Finally, changes in the PTM and the protein-level abundances are combined to remove bias due to the confounding of these two sources.

We evaluated the proposed framework on datasets from computer simulations, benchmark controlled mixtures, and biological investigations \todo{expand on the relevance and the comprehensive nature of the datasets - diverse organisms, acquisitions}. The evaluations demonstrated that by appropriately leveraging the information from \todo{which sources}, the proposed approach improves the accuracy of the estimates of PTM fold changes and results in a better calibrated proportion of false positive \todo{what?}. \todo{In particular, accounting for the confounding from protein abundance...}

The proposed approach is implemented as a freely available open source R package $MSstatsPTM$, available on Bioconductor, which employs similar input format as in $MSstats$ and $MSstatsTMT$, and is compatible with many acquisition methods, such as label free, DIA, DDA, and TMT\cite{Choi:2014,Huang:2020}. \todo{we claim that the results are reproducible - what does this mean?}

@@

The datasets were chosen to provide a performance benchmark of the proposed approach in situations with known ground truth, and to represent a variety of experimental designs and acquisition methods. The computer simulations varied in experimental realism. The first simulation was perfectly clean, with many replicates and no missing values, while the second simulation introduced real world qualities, such as limited modified features and missing values. The spike-in experiment took the real world qualities a step further and allowed us to compare the methods in a real experiment with known \todo{changes in what?}. Finally, three biological experiments demonstrated the applicability of the proposed approach across different biological organisms, experimental designs, and data acquisition strategies.

>>>>

This manuscript proposes a general statistical analysis framework that detects relative changes in post-translational modifications. The proposed framework combines the quantitative information pertaining to a PTM site over peptides and replicates in multiple conditions, and takes into account various sources of variation and confounding. The framework requires an experimental design that, for each sample, quantifies the peptides with post-translational modifications, as well as  peptides from the same proteins with no modification sites. It is applicable to both label-free and tandem mass tag (TMT)-based acquisitions. The proposed framework summarizes the abundances of peptides with and without the modification sites, and fits a linear mixed effects model that automatically adjusts to the biological and technological aspects of the experimental design. Finally changes in the PTM and protein-level abundances are combined to remove bias due to the confounding of these two sources.

We evaluated the proposed framework on datasets from computer simulations, benchmark controlled mixtures, and biological investigations, which illustrate a diverse set of organisms, acquisition methods, and experimental designs, showing the applicability of the methods to a variety of situations. The evaluations demonstrated that by appropriately leveraging the information from the unmodified peptide, the proposed approach improves the accuracy of the estimates of PTM fold changes and results in a better calibrated false positive rate. In particular, accounting for the confounding from protein abundance allows us to see the true effect of the modification, avoiding the need for more manual and time intensive methods.

The proposed approach is implemented as a freely available open source R package $MSstatsPTM$, available on Bioconductor, which employs similar input format as in $MSstats$ and $MSstatsTMT$, and is compatible with many acquisition methods, such as label free, DIA, DDA, and TMT\cite{Choi:2014,Huang:2020}.

@@

The datasets were chosen to provide a performance benchmark of the proposed approach in situations with known ground truth, and to represent a variety of experimental designs and acquisition methods. The computer simulations varied in experimental realism. The first simulation was perfectly clean, with many replicates and no missing values, while the second simulation introduced real world qualities, such as limited modified features and missing values. The spike-in experiment took the real world qualities a step further and allowed us to compare the methods in a real experiment with known changes in modified spike-in peptides. Finally, three biological experiments demonstrated the applicability of the proposed approach across different biological organisms, experimental designs, and data acquisition strategies. A summary of the experiments used in this manuscript can be seen in \todo{add and reference table}.
